---
title: "Effects of age and gender on using gene alteration and expression markers as prognostic markers for diffuse large B cell lymphoma (DLBCL)"
author: "cyy36@cam.ac.uk"
date: "6/11/2025"
output:
  prettydoc::html_pretty:
    theme: architect
    toc: yes
bibliography: references.bib
---

<center>
![](https://onlinecareeraccelerators.pace.cam.ac.uk/hubfs/CAM%20DS%20PACE/Professional%20and%20Continuing%20Education_V_reversed%20colour%201.svg)
</center>


<h3 style="text-align: center;">MSt Healthcare Data Science</h2>

#### Authentication of practice

|   |   |
|---|---|
|I confirm that I have fully read and understood the [assignment brief](https://vle.pace.cam.ac.uk/mod/page/view.php?id=2419694) for this module.| **Y** |

#### Details

|   |   |
|---|---|
|Name: Chung Yan|Surname: Yu|
|Submission Date| 19-10-25|
|Word Count: whole assignment including codes| |
|Word Count: main body excluding abstract, references and supplementary materials| |

#### Permission to share your assignment: delete as appropriate

I do give permission to share my assignment with future MSt participants.

#### University statement of originality

This assignment is the result of my own work and includes nothing which is the outcome of work done in collaboration except as declared in the Preface and specified in the text. It is not substantially the same as any that I have previously submitted for a degree or diploma or other qualification at the University of Cambridge or any other university or similar institution, or that is being concurrently submitted, except as declared in the Preface and specific in the text. I further state that no substantial part of my Portfolio has already been submitted, nor is being concurrently submitted for any such degree, diploma or other qualification at the University of Cambridge or any other university or similar institution except as declared in the Preface and specified in the text.

|   |   |
|---|---|
|I confirm the statement of originality as above| **Y** |


#### Questions for reflection

Self-assessment is an important aspect of feedback literacy, which is, in turn, key to the development of expertise. As you proceed through the MSt Healthcare data science programme, we hope that you will make use of the following prompts to assess your own work on assignments. Specific assignment briefs will likely indicate which of these to address for which assessments, but, in general, we expect you to respond to one or two for each assignment on your course. 

For each of the questions, do not spend too long answering – keep it brief. For each question you answer, limit yourself to no more than three items. And please remember, this is optional and developmental: these cover sheets are designed to create space for self-assessment and feedback dialogue, rather than additional assignment workload.

1. Which aspects of this assignment are you most uncertain about and/or would most like to receive feedback on?
2. What elements are you left pondering after this assignment that you would like to discuss further?
3. How have you incorporated feedback from peers and tutors into this assignment?
4. How, and to what extent, have you been able to incorporate feedback on previous course work into this assignment? 
5. Using the wording in the rubric, how would you describe the quality of the different aspects of your work?

### Declaration of the use of generative AI

|   |   |
|---|---|
|Which permitted use of generative AI are you acknowledging?|Semantic search of literature and notes, outline creation, code debugging, informed feedback|
|Which generative AI tool did you use (name and version)?|Claude Code v2.0.3x (the version likely changed over usage), Claude Desktop v1.0.3x with PubMed Connector, M365 CoPilot|
|What did you use the tool for?|Searching for relevant journal publications, sanity check on coding strategies, collating notes in my Obsidian vault, debugging code by parsing error messages|
|How have you used or changed the generative AI's output|E.g. my collated notes always stay in point form so that I write out the paragraphs in my own words. Feedback provided by the GenAI models are weighed and assessed before being acted on.|

# Abstract

Diffuse large B cell lymphoma (DLBCL) is the most common non-Hodgkin lymphoma, with prognosis influenced by diverse factors ranging from demographic features (age, gender) to integrated scoring systems such as the International Prognostic Index (IPI), and genetic mutations altering gene expression levels. Oncogenes MYC, BCL2 and BCL6 emerged with strong associations with DLBCL, yet how demographic factors interact with these genetic drivers remains understudied. This study analysed clinical and genomic data from 1001 DLBCL patients to investigate whether age and gender modify the prognostic value of these established oncogenes.

```{r setup, include=FALSE}
# Set chunk options: hide all code, outputs, messages, warnings, and errors
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      results = 'hide', error = FALSE)

# Suppress all output during package installation and loading
suppressMessages(suppressWarnings({

  # Package installation (silent)
  if(!require(tidyverse, quietly = TRUE)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
  if(!require(knitr, quietly = TRUE)) install.packages("knitr", repos = "http://cran.us.r-project.org")
  if(!require(DT, quietly = TRUE)) install.packages("DT", repos = "http://cran.us.r-project.org")
  if(!require(readxl, quietly = TRUE)) install.packages("readxl", repos = "http://cran.us.r-project.org")
  if(!require(httr, quietly = TRUE)) install.packages("httr", repos = "http://cran.us.r-project.org")
  if(!require(here, quietly = TRUE)) install.packages("here", repos = "http://cran.us.r-project.org")
  if(!require(kableExtra, quietly = TRUE)) install.packages("kableExtra", repos = "http://cran.us.r-project.org")

  # Load packages (silent)
  library(tidyverse)
  library(knitr)
  library(DT)
  library(readxl)
  library(httr)
  library(here)
  library(kableExtra)

}))
```

# Introduction

```{r data-loading, cache = TRUE}
# Silent data loading with fallback mechanism
# Data source tracking for inline text reference

# Define data source URLs
# Default URL, direct from ScienceDirect
direct_url <- "https://ars.els-cdn.com/content/image/1-s2.0-S0092867417311212-mmc1.xlsx"
# Backup URL, from this GitHub repository
repo_url <- "https://raw.githubusercontent.com/joelyu/HDS-W1/refs/heads/main/data/mmc1-clinicalInformation.csv"

# Initialize variables for tracking data source
clinical_data <- NULL
data_source_type <- ""  # For storing inline text reference on which URL was used

# Suppress all output during data loading
suppressMessages(suppressWarnings({

  # Attempt 1: Load from original ScienceDirect URL
  tryCatch({
    temp_file <- tempfile(fileext = ".xlsx")
    download.file(direct_url, temp_file, mode = "wb", quiet = TRUE)

    # Verify file downloaded and is readable
    if (file.exists(temp_file) && file.size(temp_file) > 0) {
      clinical_data <- read_excel(temp_file, sheet = "Clinical Information", skip = 3)

      # Verify data loaded properly, i.e. matches known dimensions
      if (nrow(clinical_data) == 1001 && ncol(clinical_data) == 35) {
        data_source_type <- "original"
      } else {
        stop("Wrong dimensions in the ScienceDirect data") # Corresponds to the dimension check
      }
    } else {
      stop("download.file failed") # Corresponds to the file exist and file size != 0 check
    }

  }, error = function(e) {
    clinical_data <<- NULL  # Ensure NULL for fallback
  })

  # Attempt 2: Fallback to processed backup dataset
  if (is.null(clinical_data)) {
    tryCatch({
      clinical_data <- read_csv(repo_url, show_col_types = FALSE)

      # Verify backup data loaded properly, the same dimension checking as above
      if (nrow(clinical_data) == 1001 && ncol(clinical_data) == 35) {
        data_source_type <- "backup"
      } else {
        stop("Wrong dimensions in the GitHub backup data")
      }

    }, error = function(e) {
      clinical_data <<- NULL  # Ensure NULL for next fallback
    })
  }

  # Attempt 3: Check for local CSV file in data/ folder
  if (is.null(clinical_data)) {
    tryCatch({
      local_file_path <- here("data", "mmc1-clinicalInformation.csv")

      if (file.exists(local_file_path)) {
        clinical_data <- read_csv(local_file_path, show_col_types = FALSE)

        # Verify local data loaded properly, same dimension checking
        if (nrow(clinical_data) == 1001 && ncol(clinical_data) == 35) {
          data_source_type <- "local"
        } else {
          stop("Wrong dimensions in local data file")
        }
      } else {
        stop("Local data file not found")
      }

    }, error = function(e) {
      # If all three attempts fail, create empty dataset to prevent knitting errors
      clinical_data <<- data.frame()
      data_source_type <<- "all-failed"
    })
  }

}))

# Create inline text variables for markdown reference
if (data_source_type == "original") {
  data_source_text <- "The dataset was loaded directly from ScienceDirect's [S1 Table](https://ars.els-cdn.com/content/image/1-s2.0-S0092867417311212-mmc1.xlsx)"
} else if (data_source_type == "backup") {
  data_source_text <- "The dataset from the ScienceDirect link was unavailable therefore the processed data from [this coursework's repository](https://github.com/joelyu/HDS-W1/blob/main/data/mmc1-clinicalInformation.csv) was used instead"
} else if (data_source_type == "local") {
  data_source_text <- "Both online sources were unavailable, so the dataset was loaded from a local copy in the `data/` folder"
} else if (data_source_type == "all-failed"){
  data_source_text <- "All data sources (ScienceDirect, GitHub repository, and local file) are unavailable. Please download the CSV from [this coursework's repository](https://github.com/joelyu/HDS-W1/blob/main/data/mmc1-clinicalInformation.csv) and place it in the `data/` folder as `mmc1-clinicalInformation.csv` and knit the HTML document again so that the analysis code could be run."
}

# Store row and column counts for inline reference
if (data_source_type == "all-failed") {
  # Hardcode expected values if all data sources fail
  n_patients <- 1001
  n_variables <- 35
} else {
  n_patients <- ifelse(is.null(clinical_data), 0, nrow(clinical_data))
  n_variables <- ifelse(is.null(clinical_data), 0, ncol(clinical_data))
}
```

Diffuse large B cell lymphoma (DLBCL) is the most common type of aggressive non-Hodgkin's lymphoma (NHL) [@blood1997; @WANG2023], with a wide range of factors contributing to prognosis. For a comparable and more comprehensive pre-treatment prognosis model, the international prognostic index (IPI) emerged as a standardised model for DLBCL and other aggressive NHLs [@shipp1993], combining 5 factors: age, Ann Arbor stage classification, ECOG Performance Status, serum lactate dehydrogenase level and the number of extranodal sites of disease. By pooling factors together, the IPI outperformed simple disease stage classification models such as Ann Arbor. As prognosis models improved, so did treatment. The first generation chemotherapy of cyclophosphamide, doxorubicin, vincristine, and prednisone (CHOP) had a cure rate of 30-35% [@Fisher1993CHOP]. With the addition of rituximab to the regiment (R-CHOP) and other advances, DLBCL becomes curable for more than 60% of patients [@johnson2012doublehit]. 

While demographic factors such as age is already featured in IPI, others such as gender is not, and there is an established body of evidence to show the involvment of gender. From how males have higher hazard risks [@Carella2013male] to gender-associated pharmacokinetic differences in rituximab leading to worse treatment responses in males [@Habermann2014sexAge]. Beyond demographic and clinical factors, a series of genetic alterations came out as strong prognostic markers: BCL2 [@Gascoyne1997BCL2], BCL6 [@LoCoco1994BCL6] and MYC [@ChenevixTrench1986MYC]. These changes are observed at both the genotypic and phenotypic levels, as translocation mutations and protein overexpression respectively [@Petrich2014MYCReview]. Furthermore, patients with varying combinations of these alterations have shown worse responses to R-CHOP treatments.

With the advent of machine learning methods came attempts to construct models based on clinical and genomic data. @reddy2017paper set out achieving this by first identifying likely genomic drivers for DLBCL, which included BCL2, BCL6, MYC as well as other candidates. The genotypic alteration data was further combined with phenotypic expression data along with cell-of-origin data to create a genomic risk model, outperforming IPI in terms of accuracy. Given the breadth of @reddy2017paper's work, the effects of gender and age on the prognostic powers of these markers were unfortunately understudied. Given the importance of these two factors are already well established in other studies and cohorts, it is the goal of this study to investigate  

Here the same dataset from the same genomic risk model study by @reddy2017paper is utilised, as their [Supplementary Table S1](https://ars.els-cdn.com/content/image/1-s2.0-S0092867417311212-mmc1.xlsx). The dataset [@reddy2017dataset] comprises of `r n_patients` DLBCL patients with `r n_variables` variables including demographic, clinical and genomic data. Key variables for this study include: age (continuous), gender (male/female), gene alteration information of MYC, BCL2 and BCL6, gene expression levels of MYC, BCL2 and BCL6, response to initial treatment, and overall survival outcomes.

# Methods

`r data_source_text`

# Results

# Conclusion

# Discussion

# Conclusion

# References

::: {#refs}
:::